Prehensile Manipulation Planning: Modelling, Algorithms and Implementation

### 2021-05-21 hpp.tex
Cite RBPRM.
Add references to Brock and Vendittelli.
Put paper in format IEEE-TRO.

Robots in industrial manufacturing are today mostly programmed by
hand. They repeat the same motion thousands of times with a high
accuracy. Automating a task with some variability is however very
challenging since it requires more programming effort to integrate
sensors and motion planning in the process. The work described in this
paper is a small step to the simplification of industrial process
automation in the presence of some variability, like the variation of
the initial position of some object or the presence of unknown
obstacles. The work only covers the motion planning -- and
more accurately the manipulation planning -- issue. The integration in a whole
process is still under development. We think that it
is important not only to develop algorithms, but also to provide them
within an open-source software platform in order to make the
evaluation and then the integration of those algorithms easier.

This \paper therefore describes a software platform called \href{https://humanoid-path-planner.github.io/hpp-doc}{Humanoid Path
  Planner} taylored for manipulation planning in robotics. It can handle many types of robots, from manipulator arms to legged humanoid robots.
The main contributions are:
\begin{itemize}
\item an original and general modelling of prehensile manipulation based on
  non-linear constraints,
\item an original solver for non-linear constraints that is able to handle implicit and explicit constraints,
\item a manipulation planning algorithm that is able to tackle a great variety of manipulation planning problems,
\item an open-source software suite that implements all the above following state of the art development tools and methods.
\item a docker image of the above software with installation instructions provided with this paper.
\end{itemize}

This paper extends the work presented in \cite{mirabel-iros2016} and \cite{MirLam2018} with some new material.

@@

Manipulation planning is a particular instance of path planning were
some objects are moved by some robots.
Although several instances of
manipulation problem exist like manipulation by
pushing~\cite{BenRiv1998}, or by throwing~\cite{WooZacLyn2017}, as well as
multicontact planning \cite{Bretl2006}, \cite{LenVaiYosKhe2013},
\cite{TonDelPetParManMan2018}, in this paper, we are only concerned with
prehensile manipulation.
The configuration space of the whole system is constrained by
non-linear constraints due to the fact that objects cannot move by
themselves and should stay in a stable pose when not grasped by a
robot.  The accessible configuration space is thus a union of
submanifolds as defined in the previous section. Each of these
manifolds may moreover be a foliation where each leaf corresponds to a
stable position of an object or to a grasp of an object by a
gripper. The geometrical structure of the problem has been well
understood for a long time~\cite{AlaSimLau89}.

@@

Rearrangement planning \cite{ota2004rearrangement}, \cite{LIS210},
\cite{KroBek2015}, \cite{LerPha2015} consists in finding a sequence of
manipulation paths that move some objects from an initial pose to a
final pose. The final configuration of the robot is not specified. A
simplifying asumption is the existence of a monotone solution, that is
a solution where each object is grasped at most once and is moved from
its initial pose to its final pose \cite{stilman2008planning},
\cite{stilman2007manipulation}, \cite{srivastavaetal2014icra}, \cite{nieuwenhuisen2008effective},
\cite{ota2004rearrangement}. They mostly use two-level methods
composed of a symbolic or task planner and of a motion
planner~\cite{cambon:hal-01976081}, \cite{KaeLoz2013}, \cite{18-toussaint-RSS}.

Other contributions in manipulation planning explicitely address the problem of multi-arm manipulation \cite{GhaCorSim2009}, \cite{HarTsuLau2014}, \cite{DobBek2015}, \cite{XiaLerPha2017}.

@@

\section{Preliminaries}\label{partie:preliminary}
\subsection{Kinematic chains and Lie groups}

A kinematic chain is commonly understood as a set of rigid-body links connected
to each other by joints. Each joint has one degree of freedom either in
rotation or in translation. A configuration of the kinematic chain is represented by a vector. Each component of the vector represents the angular or linear value of the correspondig joint.

This representation although well suited for fixed base manipulator
arms is not well adapted for robots with a mobile base like wheeled
mobile robots, aerial robots or legged robots, since the mobility of the base
cannot be correctly represented by translation or rotation joints.
Representing a freeflying object by three virtual translations followed by
three virtual rotations called roll, pitch and yaw is indeed a poor workaround
due to the presence of singularities. A good illustration of this is the gimball
lock issue that arose during Apollo~13 flight. To avoid singularities, we
propose the following defintion.

@@

\paragraph{Geometric interpretations}
\begin{itemize}
\item $\reals$ (and by trivial generalization $\reals^n$): the neutral element is $0$. The tangent space at 0 is isomorphic to $\reals$ and
  $$
  \forall \theta\in\reals,\  \exp(\theta) = \theta.
  $$
\item $SE(3)$: an element $g$ of $SE(3)$ can be seen as the position of a moving frame in a fixed reference frame. A point $\mathbf{x}\in\reals^3$ is mapped to $g(\mathbf{x})$. Note that $\mathbf{x}$ is also the coordinate vector of $g(\mathbf{x})$ in the moving frame $g$. If $\mathbf{v}$, $\omega$ are a linear and an angular velocities at the origin, $(\mathbf{v},\omega)$ is \textit{transported} to $g$
  as the same linear and angular velocities expressed in the moving frame. In other words, if
  \begin{equation}\label{eq:homogeneous matrix}
  M=\left(\begin{array}{ll} R & \mathbf{t}\\ 0 & 1\end{array}\right)
    \end{equation}
  with $R\in SO(3)$ and $\mathbf{t}\in\reals^3$  is the homogeneous matrix
  representing $g$, and $(\mathbf{v},\mathbf{\omega})$ is a velocity in $T_{I_3}SE(3)$, the velocity \textit{transported} to $g$ corresponds to linear and angular velocities $R\mathbf{v}$ and $R\omega$ of the moving frame.
  Integral curves of the vector field mentioned in item 2 above correspond to helicoidal motions of constant velocity expressed in the moving frame.
\end{itemize}
$SE(2)$, $SO(3)$, and $SO(2)$ are (or are isomorphic to) subgroups of $SE(3)$ and follow the same geometrical interpretation.

@@

Following a constant velocity $\dconf\in T_{\neutral}\liegroup$ starting from $\conf_0\in\liegroup$, leads to
$$
\conf_1 = q_0.\exp(\dconf).
$$
Note that if $\liegroup=\reals$, we write
$$
\conf_1 = \conf_0 + \dconf,
$$
since the Lie group operator of $\reals$ is $+$ and $\exp_{\reals}$ is the identity.
In order to homogeneize notation, we define the following operators. For any
$\conf_0,\conf_1\in\liegroup$ and $\dconf\in T_{\neutral}\liegroup$:
\begin{eqnarray}\label{eq:plus}
  \conf_0 \oplus \dconf &\triangleq& \conf_0.\exp(\dconf) \in \liegroup,\\
  \label{eq:minus}
  \conf_1 \ominus \conf_0 &\triangleq& \log(\conf_0^{-1}.\conf_1) \in T_{\neutral}\liegroup.
\end{eqnarray}

@@

Given a kinematic chain with joints $(J_1, \cdots, J_{njoints})$, ordered in such
a way that each joint has an index bigger than its parent in the tree, the configuration space of the robot is the cartesian product of the joint configuration spaces.
$$
\CS \triangleq \CS_{J_1}\times \cdots \times \CS_{J_{njoints}}.
$$
$\CS$ naturally inherits the Lie group structure of the joint configuration spaces through the Cartesian product.
We denote by $nq_i$, $nv_i$ the sizes of the configuration and velocity vector representations of joint $J_i$, as defined in Table~\ref{tab:vector representation}.
The configuration and velocity of the robot can thus be represented by vectors of size $nq$ and $nv$ such that
$$
nq = \sum_{i=1}^{njoints} nq_i, \ \ \ nv = \sum_{i=1}^{njoints} nv_i
$$
We denote by $iq_i$, and $iv_i$ the indices of joint $i$ in the robot configuration and velocity vectors.
$$
iq_i = \sum_{j=1}^{i-1} nq_j\ \ \ \ iv_i = \sum_{j=1}^{i-1} nv_j
$$
With these definitions and notation, the linear interpolation between two robot
configurations $\conf_0$ and $\conf_1$ is naturally written:
$$
\conf(t) = \conf_0 \oplus t (\conf_1 - \conf_0)
$$
This formula generalizes the linear interpolation to robots with freeflying bases, getting rid of singularities of roll -- pitch -- yaw parameterization.
%% begin impl
Cartesian products of Lie groups are represented by Class \href{https://gepettoweb.laas.fr/hpp/hpp-pinocchio/doxygen-html/classhpp_1_1pinocchio_1_1LiegroupSpace.html}{\texttt{LiegroupSpace}}. Elements of these spaces are represented by classes
\begin{itemize}
\item \href{https://gepettoweb.laas.fr/hpp/hpp-pinocchio/doxygen-html/classhpp_1_1pinocchio_1_1LiegroupElementBase.html}{\texttt{LiegroupElement}}
\item \href{https://gepettoweb.laas.fr/hpp/hpp-pinocchio/doxygen-html/classhpp_1_1pinocchio_1_1LiegroupElementBase.html}{\texttt{LiegroupElementRef}}, and
\item \href{https://gepettoweb.laas.fr/hpp/hpp-pinocchio/doxygen-html/classhpp_1_1pinocchio_1_1LiegroupElementBase.html}{\texttt{LiegroupElementConstRef}}.
  \end{itemize}
%% end impl

@@

In this \paper, we will make use of the term Jacobian in a generalized way.
If $h$ is a differentiable function from a Lie group $\liegroup_1$ to a Lie group $\liegroup_2$, and $\conf_1$ an element of $\liegroup_1$, we will denote by $\frac{\partial h}{\partial\conf}(\conf_1)$ the operator that maps velocities in $T_{\conf_1}\liegroup_1$ to the velocity in $T_{h(\conf_1)}\liegroup_2$ \textit{transported by} $h$\footnote{If $\dconf\in T_{\conf_1}\liegroup_1$ is a velocity along a time parameteriezd curve $\gamma$, $\frac{\partial h}{\partial\conf}(\conf_1)\dconf$ is the velocity at
  the same time along Curve $h(\gamma)$.}.

@@

It is sometimes useful to produce a configuration $\conf$ that satisfies a constraint (or a set of constraints) of type~(\ref{eq:constraint}) from a configuration $\conf_0$ that does not. This action is called the \textit{projection of $\conf_0$ onto the submanifold defined by the constraint} and is performed by a Gauss-Newton solver~\cite[Chapter~10]{NoceWrig06} that iteratively linearizes the constraint as follows:
$$
h(\conf_{i+1}) \approx h(\conf_i) + \frac{\partial h}{\partial \conf}(\conf_{i})(\conf_{i+1}-\conf_i) = 0
$$
Iterate $\conf_{i+1}$ is computed as follows:
\begin{equation}\label{eq:HierarchicalIterative}
\conf_{i+1} = \conf_i-\alpha_i\frac{\partial h}{\partial \conf}^{+}(\conf_{i}) h(\conf_i)
\end{equation}
where $.^{+}$ denotes the Moore Penrose\footnote{He has just been awarded the Nobel Prize.} pseudo inverse, and $\alpha_i$ is a positive real number called the step size. Taking $\alpha_i=1$ exactly solves the linear approximation, but it may not be the best choice in general.

@@

In manipulation planning applications, where robots manipulate objects, once an object is grasped, the position of the object can be computed explicitely from the configuration of the robot. In this case, some configuration variables of the system depend on other:
$$
\conf=(\conf_{rob},\conf_{obj})\in\CS,\ \ \ \ \conf_{obj} = g_{grasp}(\conf_{rob}).
$$
Although this constraint may fit definition~(\ref{eq:constraint}) by defining
\begin{equation}\label{eq:explicit-grasp}
h(\conf) \triangleq \conf_{obj} \ominus g_{grasp}(\conf_{rob}),
\end{equation}
solving this constraint possibly with other constraints using an iterative scheme~(\ref{eq:HierarchicalIterative}) is obviously sub-optimal.

@@

Now that we are able to project configurations onto submanifolds defined by numerical constraints -- up to some numerical threshold, we need now to define paths on such submanifolds. The usual way of doing so is to discretize the path and to project each sample configuration. The shortcoming of this method is that is requires to choose a discretization step at path construction and to lose the continuous information of the path.

Instead, we propose an alternative architecture where paths store the constraints they are subject to and apply the constraints at path evaluation. Let $P\in C^1([0,T],\CS)$ be a path without constraint defined on an interval $[0,T]$, and $\mathbf{proj}$ a projector onto a submanifold defined by numerical constraints (\textit{i.e.} an instance of\\ \href{https://gepettoweb.laas.fr/hpp/hpp-constraints/doxygen-html/classhpp_1_1constraints_1_1solver_1_1BySubstitution.html}{\texttt{SolverBySubstitution}}).

@@

The previous sections have presented how we model kinematic chains, configurations and velocities for a given robotic system, how configurations and paths can be projected onto a submanifold of the configuration space defined by numerical constraints.

@@

\paragraph{Geometric interpretation and examples}
The 3 first components of $\bar{h}(\conf)$ in equation~(\ref{eq:barh}) correspond to the position of the center of $\handle(\conf)$ in the frame of $\gripper(\conf)$. The 3 last components of $\bar{h}(\conf)$ is a vector representing the relative orientation of $\handle(\conf)$ with respect to $\gripper(\conf)$. The direction of the vector represents the axis of rotation, the norm of the vector represents the angle of rotation.
\begin{itemize}
\item If $flags=(\true,\true,\true,\true,\true,\true)$ the grasp is satisfied iff $\gripper(\conf)$ and $\handle(\conf)$ coincide:
  $$ h =\bar{h},$$
  $h_{comp}$ is an empty constraint;
\item if $flags=(\true,\true,\true,\true,\true,\false)$ the grasp is satisfied iff the centers and $z$ axes of $\gripper(\conf)$ and $\handle(\conf)$ coincide (free rotation around $z$). This is useful for cylindical objects:
  \begin{align*}
    h &= (\bar{h}_1,\bar{h}_2,\bar{h}_3,\bar{h}_4,\bar{h}_5),\\
    h_{comp} &= (\bar{h}_6);
  \end{align*}
\item if $flags=(\true,\true,\true,\false,\false,$ $\false)$ the grasp is satisfied iff the centers of $\gripper(\conf)$ and $\handle(\conf)$ coincide (free rotation). This is useful for spherical objects.
  \begin{align*}
    h &= (\bar{h}_1,\bar{h}_2,\bar{h}_3),\\
    h_{comp} &= (\bar{h}_4,\bar{h}_5,\bar{h}_6).
  \end{align*}
\end{itemize}
If $\conf_0$ is a configuration satisfying the grasp constraint: $h(\conf_0)$=0, then the submanifold defined by
$$
\left\{\conf\in\CS,\ \ h(\conf)=0\ \ h_{comp}(\conf) =  h_{comp}(\conf_0)\right\}
$$
contains all the configurations that are reachable from $\conf_0$ while keeping the grasp. Note that this representation of relative pose constraints has been used in the Stack of Task software, although it is not described in the corresponding paper~\cite{SoT}. It is different from Task Space Regions \cite{berenson2011} where open domains of $SE(3)$ are defined.

@@

When an object is not grasped, it should lie in a stable pose. There are two simple methods to enforce that:
\begin{enumerate}
\item defining virtual grippers in the environment and virtual handles on the object, implicitely defines a discrete set of poses,
\item defining a virtual gripper on an horizontal plane and a virtual handle on the object, and using a grasp with flags $(\false,\false,\true,\true,\true,\false)$ constrains the object to move on a infinite horizontal plane.
\end{enumerate}

@@

We denote by $o_i(\conf)$ and $f_j(\conf)$ the poses in configuration $\conf$ of frames with respective centers $C_{o_i}$ and $C_{f_j}$ and with $x$-axis normal to each polygon. In an analogous way as in Definition~\ref{def:grasp}, we define
\begin{equation}\label{eq:convex-shape-contact-hbar}
\bar{h}(\conf) = \log_{\reals^3\times SO(3)} \left(f_{j}(\conf)^{-1}o_i(\conf)\right)
\end{equation}
The contact constraint is defined by the following piecewise differentiable function:
\begin{equation}\label{eq:convex-shape-contact}
h(\conf) = \left\{\begin{array}{lll}(\bar{h}_1,0,0,\bar{h}_4,\bar{h}_5)&
\mbox{if} & P(C_{o_i},f_j)\mbox{ inside } f_j\\
(\bar{h}_1,\bar{h}_2,\bar{h}_3,\bar{h}_4,\bar{h}_5)&
\mbox{if} & P(C_{o_i},f_j)\mbox{ outside } f_j\end{array}\right.
\end{equation}
It is straightforward that the above function vanishes if and only if two convex polygons $o_i$ and $f_j$ are in contact and $C_{o_i}$ is inside $f_j$.

As for grasps, we need to define a parameterized complement constraint for the contact constraint in order to specify the submanifold of configurations that are reachable from one configuration while keeping the object in a constant stable pose. The naive way consists in defining
$$
h_{comp}(\conf) = \left\{\begin{array}{lll}(\bar{h}_2,\bar{h}_3,\bar{h}_6)&
\mbox{if} & P(C_{o_i},f_j)\mbox{ inside } f_j\\
(0,0,\bar{h}_6)&
\mbox{if} & P(C_{o_i},f_j)\mbox{ outside } f_j\end{array}\right.
$$
$\bar{h}_2$, $\bar{h}_3$, $\bar{h}_6$ respectively represent the translation in $x-y$ plane and the rotation around $z$-axis of frame $o_i$ with respect to frame $f_j$. Let $\conf_0$ be a configuration such that $h(\conf_0)=0$. The submanifold defined by
\begin{equation}\label{eq:stable-pose}
\left\{\conf\in\CS,\ \ h(\conf)=0\ \ h_{comp}(\conf) = h_{comp}(\conf_0) \right\}
\end{equation}
contains one object pose for each pair of polygons $(o_i,f_j)$, and there are
$|I| . |J|$ possible combinations. This constraint is therefore not suitable to
enforce object immobility along a path since the object may jump from one pose to another one.

>>>>

Robots in industrial manufacturing are today mostly programmed by
hand. They repeat the same motion thousands of times with a high
accuracy. Automating a task with some variability is however very
challenging since it requires more programming effort to integrate
sensors and motion planning in the process. A good example of this difficulty is
the Amazon picking challenge \cite{brock-2018}. The work described in this
paper is a small step to the simplification of industrial process
automation in the presence of some variability, like the variation of
the initial position of some object or the presence of unknown
obstacles. The work only covers the motion planning -- and
more accurately the manipulation planning -- issue. The integration in a whole
process is still under development. We think that it
is important not only to develop algorithms, but also to provide them
within an open-source software platform in order to make the
evaluation and then the integration of those algorithms easier.

This \paper therefore describes a software platform called \href{https://humanoid-path-planner.github.io/hpp-doc}{Humanoid Path
  Planner} taylored for manipulation planning in robotics. It can handle many types of robots, from manipulator arms to legged humanoid robots.
The main contributions are:
\begin{itemize}
\item an original and general modeling of prehensile manipulation based on
  non-linear constraints,
\item an original solver for non-linear constraints that is able to handle implicit and explicit constraints,
\item a manipulation planning algorithm that is able to tackle a great variety of manipulation planning problems,
\item an open-source software suite that implements all the above following state of the art development tools and methods.
\item a docker image of the above software with installation instructions provided with this paper. This image makes the experimental results replicable.
\end{itemize}

This paper extends the work presented in \cite{mirabel-iros2016} and \cite{MirLam2018} with the following new material:
\begin{itemize}
\item description of the configuration space as a Cartesian product of Lie groups (Section~\ref{partie:preliminary}),
\item unified and detailed definition of the grasp and placement constraints that are only evoked in \cite{mirabel-iros2016} (Section~\ref{sec:manipulation-problem}),
\item automatic construction of the constraint graph (Section~\ref{sec:manipulation-problem}),
\item the docker image of the software,
\item a description of the software platform (Section~\ref{sec:humanoid-path-planner}),
\item experimental results on several different problems.
\end{itemize}

@@

Manipulation planning is a particular instance of path planning were
some objects are moved by some robots.
Although several instances of
manipulation problem exist like manipulation by
pushing~\cite{BenRiv1998}, or by throwing~\cite{WooZacLyn2017}, as well as
multi-contact planning \cite{Bretl2006}, \cite{LenVaiYosKhe2013},
\cite{TonDelPetParManMan2018}, in this paper, we are only concerned with
prehensile manipulation.
The configuration space of the whole system is constrained by
non-linear constraints due to the fact that objects cannot move by
themselves and should stay in a stable pose when not grasped by a
robot.  The accessible configuration space is thus a union of
sub-manifolds as defined in the previous section. Each of these
manifolds may moreover be a foliation where each leaf corresponds to a
stable position of an object or to a grasp of an object by a
gripper. The geometrical structure of the problem has been well
understood for a long time~\cite{AlaSimLau89}. Some specific instances of the
problem have even been addressed recently \cite{vendittelli-2018}.

@@

Rearrangement planning \cite{ota2004rearrangement}, \cite{LIS210},
\cite{KroBek2015}, \cite{LerPha2015} consists in finding a sequence of
manipulation paths that move some objects from an initial pose to a
final pose. The final configuration of the robot is not specified. A
simplifying assumption is the existence of a monotone solution, that is
a solution where each object is grasped at most once and is moved from
its initial pose to its final pose \cite{stilman2008planning},
\cite{stilman2007manipulation}, \cite{srivastavaetal2014icra}, \cite{nieuwenhuisen2008effective},
\cite{ota2004rearrangement}. They mostly use two-level methods
composed of a symbolic or task planner and of a motion
planner~\cite{cambon:hal-01976081}, \cite{KaeLoz2013}, \cite{18-toussaint-RSS}.

Other contributions in manipulation planning explicitly address the problem of multi-arm manipulation \cite{GhaCorSim2009}, \cite{HarTsuLau2014}, \cite{DobBek2015}, \cite{XiaLerPha2017}.

@@

\section{Preliminaries: kinematic chains and Lie groups}\label{partie:preliminary}

A kinematic chain is commonly understood as a set of rigid-body links connected
to each other by joints. Each joint has one degree of freedom either in
rotation or in translation. A configuration of the kinematic chain is represented by a vector. Each component of the vector represents the angular or linear value of the corresponding joint.

This representation although well suited for fixed base manipulator
arms is not well adapted for robots with a mobile base like wheeled
mobile robots, aerial robots or legged robots, since the mobility of the base
cannot be correctly represented by translation or rotation joints.
Representing a free-flying object by three virtual translations followed by
three virtual rotations called roll, pitch and yaw is indeed a poor workaround
due to the presence of singularities. A good illustration of this is the gimball
lock issue that arose during Apollo~13 flight. To avoid singularities, we
propose the following definition.

@@

\paragraph{Geometric interpretations}
\begin{itemize}
\item $\reals$ (and by trivial generalization $\reals^n$): the neutral element is $0$. The tangent space at 0 is isomorphic to $\reals$ and
  $$
  \forall \theta\in\reals,\  \exp(\theta) = \theta.
  $$
\item $SE(3)$: an element $g$ of $SE(3)$ can be seen as the position of a moving frame in a fixed reference frame. A point $\mathbf{x}\in\reals^3$ is mapped to $g(\mathbf{x})$. Note that $\mathbf{x}$ is also the coordinate vector of $g(\mathbf{x})$ in the moving frame $g$. If $\mathbf{v}$, $\omega$ are a linear and an angular velocities at the origin, $(\mathbf{v},\omega)$ is \textit{transported} to $g$
  as the same linear and angular velocities expressed in the moving frame. In other words, if
  \begin{equation}\label{eq:homogeneous matrix}
  M=\left(\begin{array}{ll} R & \mathbf{t}\\ 0 & 1\end{array}\right)
    \end{equation}
  with $R\in SO(3)$ and $\mathbf{t}\in\reals^3$  is the homogeneous matrix
  representing $g$, and $(\mathbf{v},\mathbf{\omega})$ is a velocity in $T_{I_3}SE(3)$, the velocity \textit{transported} to $g$ corresponds to linear and angular velocities $R\mathbf{v}$ and $R\omega$ of the moving frame.
  Integral curves of the vector field mentioned in item 2 above correspond to screw motions of constant velocity expressed in the moving frame.
\end{itemize}
$SE(2)$, $SO(3)$, and $SO(2)$ are (or are isomorphic to) subgroups of $SE(3)$ and follow the same geometrical interpretation.

@@

Following a constant velocity $\dconf\in T_{\neutral}\liegroup$ starting from $\conf_0\in\liegroup$, leads to
$$
\conf_1 = q_0.\exp(\dconf).
$$
Note that if $\liegroup=\reals$, we write
$$
\conf_1 = \conf_0 + \dconf,
$$
since the Lie group operator of $\reals$ is $+$ and $\exp_{\reals}$ is the identity.
In order to homogenize notation, we define the following operators. For any
$\conf_0,\conf_1\in\liegroup$ and $\dconf\in T_{\neutral}\liegroup$:
\begin{eqnarray}\label{eq:plus}
  \conf_0 \oplus \dconf &\triangleq& \conf_0.\exp(\dconf) \in \liegroup,\\
  \label{eq:minus}
  \conf_1 \ominus \conf_0 &\triangleq& \log(\conf_0^{-1}.\conf_1) \in T_{\neutral}\liegroup.
\end{eqnarray}

@@

Given a kinematic chain with joints $(J_1, \cdots, J_{njoints})$, ordered in such
a way that each joint has an index bigger than its parent in the tree, the configuration space of the robot is the Cartesian product of the joint configuration spaces.
$$
\CS \triangleq \CS_{J_1}\times \cdots \times \CS_{J_{njoints}}.
$$
$\CS$ naturally inherits the Lie group structure of the joint configuration spaces through the Cartesian product.
We denote by $nq_i$, $nv_i$ the sizes of the configuration and velocity vector representations of joint $J_i$, as defined in Table~\ref{tab:vector representation}.
The configuration and velocity of the robot can thus be represented by vectors of size $nq$ and $nv$ such that
$$
nq = \sum_{i=1}^{njoints} nq_i, \ \ \ nv = \sum_{i=1}^{njoints} nv_i
$$
We denote by $iq_i$, and $iv_i$ the indices of joint $i$ in the robot configuration and velocity vectors.
$$
iq_i = \sum_{j=1}^{i-1} nq_j\ \ \ \ iv_i = \sum_{j=1}^{i-1} nv_j
$$
With these definitions and notation, the linear interpolation between two robot
configurations $\conf_0$ and $\conf_1$ is naturally written:
$$
\conf(t) = \conf_0 \oplus t (\conf_1 - \conf_0)
$$
This formula generalizes the linear interpolation to robots with free-flying bases, getting rid of singularities of roll -- pitch -- yaw parameterization.
%% begin impl
Cartesian products of Lie groups are represented by Class \href{https://gepettoweb.laas.fr/hpp/hpp-pinocchio/doxygen-html/classhpp_1_1pinocchio_1_1LiegroupSpace.html}{\texttt{LiegroupSpace}}. Elements of these spaces are represented by classes
\begin{itemize}
\item \href{https://gepettoweb.laas.fr/hpp/hpp-pinocchio/doxygen-html/classhpp_1_1pinocchio_1_1LiegroupElementBase.html}{\texttt{LiegroupElement}}
\item \href{https://gepettoweb.laas.fr/hpp/hpp-pinocchio/doxygen-html/classhpp_1_1pinocchio_1_1LiegroupElementBase.html}{\texttt{LiegroupElementRef}}, and
\item \href{https://gepettoweb.laas.fr/hpp/hpp-pinocchio/doxygen-html/classhpp_1_1pinocchio_1_1LiegroupElementBase.html}{\texttt{LiegroupElementConstRef}}.
  \end{itemize}
%% end impl

@@

In this \paper, we will make use of the term Jacobian in a generalized way.
If $h$ is a differentiable function from a Lie group $\liegroup_1$ to a Lie group $\liegroup_2$, and $\conf_1$ an element of $\liegroup_1$, we will denote by $\frac{\partial h}{\partial\conf}(\conf_1)$ the operator that maps velocities in $T_{\conf_1}\liegroup_1$ to the velocity in $T_{h(\conf_1)}\liegroup_2$ \textit{transported by} $h$\footnote{If $\dconf\in T_{\conf_1}\liegroup_1$ is a velocity along a time parameterized curve $\gamma$, $\frac{\partial h}{\partial\conf}(\conf_1)\dconf$ is the velocity at
  the same time along Curve $h(\gamma)$.}.

@@

It is sometimes useful to produce a configuration $\conf$ that satisfies a constraint (or a set of constraints) of type~(\ref{eq:constraint}) from a configuration $\conf_0$ that does not. This action is called the \textit{projection of $\conf_0$ onto the sub-manifold defined by the constraint} and is performed by a Gauss-Newton solver~\cite[Chapter~10]{NoceWrig06} that iteratively linearizes the constraint as follows:
$$
h(\conf_{i+1}) \approx h(\conf_i) + \frac{\partial h}{\partial \conf}(\conf_{i})(\conf_{i+1}-\conf_i) = 0
$$
Iterate $\conf_{i+1}$ is computed as follows:
\begin{equation}\label{eq:HierarchicalIterative}
\conf_{i+1} = \conf_i-\alpha_i\frac{\partial h}{\partial \conf}^{+}(\conf_{i}) h(\conf_i)
\end{equation}
where $.^{+}$ denotes the Moore Penrose\footnote{He has just been awarded the Nobel Prize.} pseudo inverse, and $\alpha_i$ is a positive real number called the step size. Taking $\alpha_i=1$ exactly solves the linear approximation, but it may not be the best choice in general.

@@

In manipulation planning applications, where robots manipulate objects, once an object is grasped, the position of the object can be computed explicitly from the configuration of the robot. In this case, some configuration variables of the system depend on other:
$$
\conf=(\conf_{rob},\conf_{obj})\in\CS,\ \ \ \ \conf_{obj} = g_{grasp}(\conf_{rob}).
$$
Although this constraint may fit definition~(\ref{eq:constraint}) by defining
\begin{equation}\label{eq:explicit-grasp}
h(\conf) \triangleq \conf_{obj} \ominus g_{grasp}(\conf_{rob}),
\end{equation}
solving this constraint possibly with other constraints using an iterative scheme~(\ref{eq:HierarchicalIterative}) is obviously sub-optimal.

@@

Now that we are able to project configurations onto sub-manifolds defined by numerical constraints -- up to some numerical threshold, we need now to define paths on such sub-manifolds. The usual way of doing so is to discretize the path and to project each sample configuration. The shortcoming of this method is that is requires to choose a discretization step at path construction and to lose the continuous information of the path.

Instead, we propose an alternative architecture where paths store the constraints they are subject to and apply the constraints at path evaluation. Let $P\in C^1([0,T],\CS)$ be a path without constraint defined on an interval $[0,T]$, and $\mathbf{proj}$ a projector onto a sub-manifold defined by numerical constraints (\textit{i.e.} an instance of\\ \href{https://gepettoweb.laas.fr/hpp/hpp-constraints/doxygen-html/classhpp_1_1constraints_1_1solver_1_1BySubstitution.html}{\texttt{SolverBySubstitution}}).

@@

The previous sections have presented how we model kinematic chains, configurations and velocities for a given robotic system, how configurations and paths can be projected onto a sub-manifold of the configuration space defined by numerical constraints.

@@

\paragraph{Geometric interpretation and examples}
The 3 first components of $\bar{h}(\conf)$ in equation~(\ref{eq:barh}) correspond to the position of the center of $\handle(\conf)$ in the frame of $\gripper(\conf)$. The 3 last components of $\bar{h}(\conf)$ is a vector representing the relative orientation of $\handle(\conf)$ with respect to $\gripper(\conf)$. The direction of the vector represents the axis of rotation, the norm of the vector represents the angle of rotation.
\begin{itemize}
\item If $flags=(\true,\true,\true,\true,\true,\true)$ the grasp is satisfied iff $\gripper(\conf)$ and $\handle(\conf)$ coincide:
  $$ h =\bar{h},$$
  $h_{comp}$ is an empty constraint;
\item if $flags=(\true,\true,\true,\true,\true,\false)$ the grasp is satisfied iff the centers and $z$ axes of $\gripper(\conf)$ and $\handle(\conf)$ coincide (free rotation around $z$). This is useful for cylindrical objects:
  \begin{align*}
    h &= (\bar{h}_1,\bar{h}_2,\bar{h}_3,\bar{h}_4,\bar{h}_5),\\
    h_{comp} &= (\bar{h}_6);
  \end{align*}
\item if $flags=(\true,\true,\true,\false,\false,$ $\false)$ the grasp is satisfied iff the centers of $\gripper(\conf)$ and $\handle(\conf)$ coincide (free rotation). This is useful for spherical objects.
  \begin{align*}
    h &= (\bar{h}_1,\bar{h}_2,\bar{h}_3),\\
    h_{comp} &= (\bar{h}_4,\bar{h}_5,\bar{h}_6).
  \end{align*}
\end{itemize}
If $\conf_0$ is a configuration satisfying the grasp constraint: $h(\conf_0)$=0, then the sub-manifold defined by
$$
\left\{\conf\in\CS,\ \ h(\conf)=0\ \ h_{comp}(\conf) =  h_{comp}(\conf_0)\right\}
$$
contains all the configurations that are reachable from $\conf_0$ while keeping the grasp. Note that this representation of relative pose constraints has been used in the Stack of Task software, although it is not described in the corresponding paper~\cite{SoT}. It is different from Task Space Regions \cite{berenson2011} where open domains of $SE(3)$ are defined.

@@

When an object is not grasped, it should lie in a stable pose. There are two simple methods to enforce that:
\begin{enumerate}
\item defining virtual grippers in the environment and virtual handles on the object, implicitly defines a discrete set of poses,
\item defining a virtual gripper on an horizontal plane and a virtual handle on the object, and using a grasp with flags $(\false,\false,\true,\true,\true,\false)$ constrains the object to move on a infinite horizontal plane.
\end{enumerate}

@@

We denote by $o_i(\conf)$ and $f_j(\conf)$ the poses in configuration $\conf$ of frames with respective centers $C_{o_i}$ and $C_{f_j}$ and with $x$-axis normal to each polygon. In an analogous way as in Definition~\ref{def:grasp}, we define
\begin{equation}\label{eq:convex-shape-contact-hbar}
\bar{h}(\conf) = \log_{\reals^3\times SO(3)} \left(f_{j}(\conf)^{-1}o_i(\conf)\right)
\end{equation}
The contact constraint is defined by the following piece-wise differentiable function:
\begin{equation}\label{eq:convex-shape-contact}
h(\conf) = \left\{\begin{array}{lll}(\bar{h}_1,0,0,\bar{h}_4,\bar{h}_5)&
\mbox{if} & P(C_{o_i},f_j)\mbox{ inside } f_j\\
(\bar{h}_1,\bar{h}_2,\bar{h}_3,\bar{h}_4,\bar{h}_5)&
\mbox{if} & P(C_{o_i},f_j)\mbox{ outside } f_j\end{array}\right.
\end{equation}
It is straightforward that the above function vanishes if and only if two convex polygons $o_i$ and $f_j$ are in contact and $C_{o_i}$ is inside $f_j$.

As for grasps, we need to define a parameterized complement constraint for the contact constraint in order to specify the sub-manifold of configurations that are reachable from one configuration while keeping the object in a constant stable pose. The naive way consists in defining
$$
h_{comp}(\conf) = \left\{\begin{array}{lll}(\bar{h}_2,\bar{h}_3,\bar{h}_6)&
\mbox{if} & P(C_{o_i},f_j)\mbox{ inside } f_j\\
(0,0,\bar{h}_6)&
\mbox{if} & P(C_{o_i},f_j)\mbox{ outside } f_j\end{array}\right.
$$
$\bar{h}_2$, $\bar{h}_3$, $\bar{h}_6$ respectively represent the translation in $x-y$ plane and the rotation around $z$-axis of frame $o_i$ with respect to frame $f_j$. Let $\conf_0$ be a configuration such that $h(\conf_0)=0$. The sub-manifold defined by
\begin{equation}\label{eq:stable-pose}
\left\{\conf\in\CS,\ \ h(\conf)=0\ \ h_{comp}(\conf) = h_{comp}(\conf_0) \right\}
\end{equation}
contains one object pose for each pair of polygons $(o_i,f_j)$, and there are
$|I| . |J|$ possible combinations. This constraint is therefore not suitable to
enforce object immobility along a path since the object may jump from one pose to another one.
